---
title: "What is Population Ecology?"
author: "Matt"
date: "1 7 2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is population ecology?

Population ecology:

The dynamics of species populations and how these populations interact with the environment

## What is a population?

* A population is a all the individuals from a single species within an ecological community

## Why do we study it?

* In population ecology we are interested in growth, fluctuations, spread, interactions (within and between populations)


* What information might we need to gather?
* How would you measure population ecology?


## Population change

* Methods to predict population change differ depending on whether reproduction only occurs seasonally, or whether it is continuous

* Geometric and exponential rates of growth

* Determining when exponential growth occurs

## Population growth

* Some species only breed once a year, with a distinct discrete breeding season.
e.g. Most invertebrates, plants, mammals
“Geometric growth”

* Other species can breed at any time, so that the populations show continuous growth
e.g. bacteria, primates
“Exponential growth”

* Different methods to predict populations

## Population growth (2)

* Imagine population of 2 animals, producing 2 offspring
* Ignore Death, Emmigration and Immigration

```{r annual breeding, echo=FALSE}
N=c(NA,NA,NA,NA,NA,NA,NA,NA,NA)
N[1]=2
N[2]=4

for (i in 2:10){
  N[i+1]<-N[i]+N[i-1]
}
dat<-data.frame(N=N, Year=as.integer(c(0:10)))
library(ggplot2)
ggplot(data=dat,(aes(Year, N)))+
  geom_point(colour="darkgreen", size=4)+
  ggthemes::theme_base()

```

## Population breeds seasonally (i.e. at discrete time intervals)

* Population growth is described using difference equations
* Difference equations show the relationship between the population size at the start compared to the population at the end of one timestep
 - Timestep is usually a year etc.

* Geometric growth if population unlimited

## Equation in words

Pop. at end of timestep = Pop. at start of timestep   x  Discrete per capita net growth rate


## Discrete breeding - unlimited population growth equation

$$Nt+1 = λ.Nt$$

Nt 	population size at time t (this year)

Nt+1	population size at time t+1 (next year)

λ		finite rate of population increase, or 	discrete (per capita) net rate of growth, 	or geometric growth rate (births-deaths)

t		is the time period (e.g. in years)

## What is λ ?

* λ (Greek symbol ‘lambda’) is the growth rate per head, and is any number greater than zero

* Note that values of λ less than 1 result in population decline, so the term ‘growth’ rate is a little confusing.

* Some textbooks use R instead of λ

 $$𝜆=  𝑁_(𝑡+1)/𝑁_𝑡 $$

## Population size after discrete (seasonal) reproduction over many years

* If you had two timesteps, then
 - Nt+2 = λ.Nt+1
 - which is the same as
 - Nt+2 = λ.λ.Nt = λ2.Nt
 - so that the generalised form for any number of years (T) is:

NT = λT .N0

## Species breeds all year (continuous reproduction)

* Population growth in these animals is usually described with differential equations

* Differential equations describe rates of change

* Exponential growth if population unlimited


## Equation in words

Rate of increase in population = (Average per capita birth rate - average per capita death rate)  x   Current population size


## Continuous breeding: unlimited population growth equation

𝑑𝑁/𝑑𝑡=(𝐵−𝐷)𝑁

𝑑𝑁/𝑑𝑡=𝑟𝑁

dN = change in population size

dt = change in time

dN/dt = rate of change in population

r = instantaneous rate of population growth


## λ versus r

* Note that λ and r are not the same!
* λ = exp(r)
 or
 r = ln(λ)
* λ only compounds once a year, whereas r compounds instantaneously
* So, for example, if λ is 1.2 then the equivalent growth rate in r is only 0.18

* Think of bank interest rates on monthly vs annual accounts

## Population size after continuous reproduction over many years

Continuous population growth after a set period of time can be predicted from:

𝑁_𝑡=𝑁_0 𝑒^𝑟𝑡

where:

N0 =  Population at start

Nt  = Population at time t

r = instantaneous population growth rate

e = Euler’s number for natural logarithm 2.718…


## Limited population growth

* In reality no population can grow indefinitely

* Values of λ or r fluctuate both independent and dependent on the density of the population

* Eventually density-dependent will limit growth; observed in a wide range of taxa and habitats

* Carrying capacity K must be considered

## Intraspecific competition: discrete breeding

Intraspecific competition can be represented by the discrete logistic growth equation

𝑁_(𝑡+1)=𝜆𝑁_𝑡 (1−𝑁_𝑡/𝐾)

where
K = carrying capacity
λ = net reproductive rate

## Example 


## Intraspecific competition:continuous breeding

Logistic growth equation for continuous growth:

𝑑𝑁/𝑑𝑡=𝑟𝑁(1−𝑁/𝐾)

where
K = carrying capacity
r = instantaneous population growth rate


## Can human carrying capacity be predicted?

* How will human population growth level off to a steady rate?

* Problem of aging population as reproductive rate slows

* Development – demand on resources by Developed and Developing countries

* Attempt by Pearl & Reed (1920) to estimate K in USA

## Example


## Inverse density dependence

* So far we have only considered populations where increasing population size results in a decline in population growth rates.

* Density dependence usually has a regulating effect, so that the population tends towards its carrying capacity.

* Inverse density dependence occurs where population growth rates fall as the population size declines.

* THIS IS DESTABILISING AND CAN DRIVE ENDANGERED POPULATION TOWARDS EXTINCTION!

## Example

## Allee effect

* Why did this happen in guillemots?  

* Predation rate (from gulls) on chicks reduced with larger colonies.

 - “Predator saturation”

* Applies to many animals.  
- Low populations: problems of finding mates, inbreeding etc.

## Examples

