---
title: "If you go down to the woods today...."
author: "Aung Ko Lin"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bear population challenge
Apply your knowledge about population models to this grizzly bear dataset. Answer the questions and show your code (either in a new script file or in a new Rmarkdown file - you can adapt the original). Save the file and send it to me. 


```{r load library}
# a # sign means that R does not read this as code. 
#install.packages("popbio") #remove the first hash if you do not have popbio installed as a package 
library(popbio)
```

## Upload and explore the bears dataset



```{r bears, echo=TRUE}
data(grizzly)
View(grizzly)
#?grizzly# what is the grizzly data about?
attach(grizzly) ## You need to attach the data so that 
#you can use it in the functions that follow 
#(otherwise you need to signal to R which data you are using)
plot(year, N, type='o', pch=16, las=1, xlab="Year", 
     ylab="Adult females", main="Yellowstone grizzly bears") 
```

## What is the grizzly data set all about

Look up the grizzly dataset from the R help (see above ?grizzly will search for the dataset in the helpfiles).

Provide a summary of the grizzly dataset
```{r}
summary(grizzly) 
```
> There are lots of other packages you could use to make a summary for example this makes a summary output (better for more complex data frames):

```{r package for summarys}
#install.packages("summarytools")
library(summarytools)
view(summarytools::dfSummary(grizzly))

```


## Calculate the rate of population change

> Good work. It might be useful (for you) to add comments to your code (use the # to write comments) so you know what you did and why

```{r}
grizzly[1:10,]
lam<-c(rep("NA",9))
for (i in 1:10){
  lam[i]<-grizzly$N[i+1]/grizzly$N[i]
}
lam
```


Their are only female bears in this count. What could you do to make this data more representative of the population?

We could do to get reliable data of other effects to represent the population of grizzly bears.

Do we need more data than just female bears?

We need more data of other effects to predict the population of grizzly bears.

> What sort of other effects are you thinking about? Expand on your answers?  Why do we need more data - does this population model answer our question? What other data could we have?

## Use the population models to assess the population change in bears 
```{r}
mean(as.numeric(lam), na.rm=TRUE)
sd(as.numeric(lam), na.rm=TRUE)
```

Make use of only the first 10 years of the bears dataset to predict the future 17 years.
Plot this on the same graph (hint in base R you can add points to a plot using the "points()" fuction)

> rather than adding in the actual number you can use the object to make the code reproducible (remember I said that your simulation will change each time you use it). So you could have rnorm(1, mean, sd) - you will have to create the objects above mean<- mean(as.numeric(lam), na.rm=TRUE) but you can not use the name mean as R uses that for itself so maybe use something like "mn"

#Extra credit:: Make the plots look pretty (look at ?ggplot2 for example and search on Google for ggplot2)
```{r}
dim(grizzly)
N=NULL

N[1]=grizzly$N[10]

for (i in  1:16){
  N[i+1]<-rnorm(1,0.9970853, 0.06100529)*N[i]
}
N
```

> Looks good. Well done

```{r}
library(tidyverse)
bears<-data.frame(N=N, years=grizzly$year[10:26])
bears %>%
  ggplot(aes(years,N))+
  geom_point(col="blue")+
  geom_line(col="blue")+
  geom_point(data=grizzly, aes(year, N), colours="brown")+ # spelling mistake here it should be "colour"
  geom_line(data=grizzly, aes(year, N), colour="brown")+
  ggthemes::theme_base()
```

```{r detach}
detach(grizzly)# remember to detach the data from the session
rm(list=ls())# this cleans your environment - is useful but 

#you might want to keep it sometimes - use caution
```


