---
title: "If you go down to the woods today...."
author: "Poldej Kochprapa (Bas)"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> I get this error when I try and knit:  Duplicate chunk label 'correction', which has been used for the chunk:

> Each chunk needs a unique name (but great that you learnt to name them!)

## Bear population challenge
Apply your knowledge about population models to this grizzly bear dataset. Answer the questions and show your code (either in a new script file or in a new Rmarkdown file - you can adapt the original). Save the file and send it to me. 


```{r load library}
# a # sign means that R does not read this as code. 
#install.packages("popbio") #remove the first hash if you do not have popbio installed as a package 
library(popbio)
```

## Upload and explore the bears dataset

> I get an error here because I have an object called N already. To avoid this in the future I need to change your plot code to call the title of the column from the data. So grizzly$year, grizzly$N

```{r bears, echo=TRUE}
data(grizzly)
View(grizzly)
#?grizzly# what is the grizzly data about?
attach(grizzly) ## You need to attach the data so that 
#you can use it in the functions that follow 
#(otherwise you need to signal to R which data you are using)
plot(grizzly$year, grizzly$N, type='o', pch=16, las=1, xlab="Year", 
     ylab="Adult females", main="Yellowstone grizzly bears") 
```

## What is the grizzly data set all about

Look up the grizzly dataset from the R help (see above ?grizzly will search for the dataset in the helpfiles).

Provide a summary of the grizzly dataset


```{r}

summary(grizzly)





```

> There are lots of other packages you could use to make a summary for example this makes a summary output (better for more complex data frames):

```{r package for summarys}
#install.packages("summarytools")
library(summarytools)
view(summarytools::dfSummary(grizzly))

```



## Calculate the rate of population change 

Their are only female bears in this count. What could you do to make this data more representative of the population?

Do we need more data than just female bears?


> You need to answer these questions



```{r calculate rate of change}

#calcualte  log(Nt+1/Nt)
nt <- length(grizzly$N) ## number transitions
logN <- log(grizzly$N[-1] / grizzly$N[-nt])
## Mean and var
c(mean = mean(logN), var = var(logN))
###
x <- sqrt(grizzly$year[-1] - grizzly$year[-length(grizzly$year)])
y <- logN / x
mod <- lm(y ~ 0 + x)
summary(mod)

N0<-44
r.mean<-0.02
r.sd<-0.02
T<-39
r<-rnorm(T,r.mean,r.sd)
plot(density(r))+
abline(v=r.mean,col="blue")+
abline(v=r.mean+r.sd,col="blue",lty=2)+
abline(v=r.mean-r.sd,col="blue",lty=2)

###
t<-N<-array(dim=T+1)

#first element is initial value
N[1]<-N0
t[1]<-0
for (i in 1:T)
{
  N[i+1]<-N[i]*exp(r[i])
  t[i+1]=t[i]+1
}

data.frame(t=t,N=N)
plot(t,N,type="l")

```



```{r correction1}
grizzly[1:10,]

lam<-c(rep("NA",9))
for (i in 1:10)
{
  lam[i]<-grizzly$N[i+1]/grizzly$N[i]
}
lam
```



## Use the population models to assess the population change in bears 

Make use of only the first 10 years of the bears dataset to predict the future 17 years.
Plot this on the same graph (hint in base R you can add points to a plot using the "points()" fuction)

```{r assess population change}
N0<-44
r.mean<-0.01
r.sd<-0.02
T<-17
r<-rnorm(T,r.mean,r.sd)
plot(density(r))+
abline(v=r.mean,col="blue")+
abline(v=r.mean+r.sd,col="blue",lty=2)+
abline(v=r.mean-r.sd,col="blue",lty=2)


```


```{r correction2}
mean<-mean(as.numeric(lam),na.rm = TRUE)
std<-sd(as.numeric(lam),na.rm = TRUE)

```

```{r correction3}
dim(grizzly)
N=NULL

N[1]=grizzly$N[10]

for (i in 1:16)
{
  N[i+1]<-rnorm(1, mean, std)*N[i]

}
N



```


```{r correction4}
library(tidyverse)
bears<-data.frame(N=N, years=grizzly$year[10:26])
bears %>%
  ggplot(aes(years,N))+
  geom_point(col="blue")+
  geom_line(col="blue")+
  geom_point(data=grizzly,aes(year,N),colour="red")+
  geom_line(data=grizzly,aes(year,N),colour="red")+
  ggthemes::theme_base()

```
#Extra credit:: Make the plots look pretty (look at ?ggplot2 for example and search on Google for ggplot2)


> Great!


```{r detach}
detach(grizzly)# remember to detach the data from the session
rm(list=ls())# this cleans your environment - is useful but 
#you might want to keep it sometimes - use caution
```


