---
title: "If you go down to the woods today...."
author: "Lay Win"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Bear population challenge
Apply your knowledge about population models to this grizzly bear dataset. Answer the questions and show your code (either in a new script file or in a new Rmarkdown file - you can adapt the original). Save the file and send it to me. 


```{r load library}
# a # sign means that R does not read this as code. 
#install.packages("popbio") #remove the first hash if you do not have popbio installed as a package 
library(popbio)
```

## Upload and explore the bears dataset

> I get an error here because I have an object called N already. To avoid this in the future I need to change your plot code to call the title of the column from the data. So grizzly$year, grizzly$N

```{r bears, echo=TRUE}
data(grizzly)
View(grizzly)
#?grizzly# what is the grizzly data about?
attach(grizzly) ## You need to attach the data so that 
#you can use it in the functions that follow 
#(otherwise you need to signal to R which data you are using)

plot(grizzly$year, grizzly$N, type='o', pch=16, las=1, xlab="Year", 
     ylab="Adult females", main="Yellowstone grizzly bears") 
```

## What is the grizzly data set all about

Look up the grizzly dataset from the R help (see above ?grizzly will search for the dataset in the helpfiles).

Provide a summary of the grizzly dataset
```{r summary}
summary(grizzly)
```

> There are lots of other packages you could use to make a summary for example this makes a summary output (better for more complex data frames):

```{r package for summarys}
#install.packages("summarytools")
library(summarytools)
view(summarytools::dfSummary(grizzly))

```

## Calculate the rate of population change 

Their are only female bears in this count. What could you do to make this data more representative of the population?

Do we need more data than just female bears?

> You need to answer these questions

## Use the population models to assess the population change in bears 

Make use of only the first 10 years of the bears dataset to predict the future 17 years.
Plot this on the same graph (hint in base R you can add points to a plot using the "points()" fuction)

#Extra credit:: Make the plots look pretty (look at ?ggplot2 for example and search on Google for ggplot2)

```{r}
grizzly[1:10,]
lam<-c(rep("NA",9))
for(i in 1:10){lam[i]<-grizzly$N[i+1]/grizzly$N[i]}
lam
```

>spaces are needed between code chunks

```{r}
mean(as.numeric(lam),na.rm=TRUE)
sd(as.numeric(lam),na.rm=TRUE)
```

 
```{r}
dim(grizzly)
N=NULL
N[1]=grizzly$N[10]
for(i in 1:16){N[i+1]<-rnorm(1,0.9970853,0.06100527)*N[i]}
N
```

> I get this error here: Error: Aesthetics must be either length 1 or the same as the data (17): x. So there is a problem with the code - I have shown how I fixed this below yours 



```{r eval=FALSE}
library(tidyverse)

bears<-data.frame(N=N,years=grizzly$year[10:26])
bears%>%
ggplot(aes(year,N))+geom_point(col="green")+geom_line(col="green")+geom_point(data=grizzly,aes(year,N),colour="blue")+geom_line(data=grizzly,aes(year,N),colour="blue")+ggthemes::theme_base()
  
```

> Check your spelling! Rememeber that an object called year is not the same as an object called years

```{r}
library(tidyverse)
bears<-data.frame(N=N,years=grizzly$year[10:26])
bears %>%
  ggplot(aes(years,N)) +
  geom_line(col="green")+
  geom_point(data=grizzly,aes(year,N),colour="blue")+
  geom_line(data=grizzly,aes(year,N),colour="blue")+
  ggthemes::theme_base()

```


> Great - you have managed to predict the population!
  


```{r detach}
detach(grizzly)# remember to detach the data from the session
rm(list=ls())# this cleans your environment - is useful but 
#you might want to keep it sometimes - use caution
```


